A Research on Big Data and AI Analysis Algorithm Optimization Using GPUs

### 2018-05-09 paper.tex
first comment from prof

In recent years, with the explosion of internet and mobile evolution, internet service is becoming bigger and bigger in terms of system complexity and user volume. Concerning user privacy and credentiality, internet service providers never want to give credential to wrong user. An obvious and canonical method is to use a combination of secret password and unique user identification. However, this becomes a burden for service users to remember their password on each service. There also is a trade-off between remembering convenience and security level when choosing a password. User behaviour of sharing password between multiple services raises the password leaking risk. Recently, there is a new way to give credentiality using Lifestyle Authentication. Lifestyle Authentication is a brand new authentication method which makes use of user auxiliary information rather than user id and password combination~\cite{weko_175884_1}. Auxiliary information can be user location paired with system access timestamp, user daily access pattern, etc\ldots. With this authentication method, a system can identify and authorize an user without requiring him/her doing any procedure explicitly. Yamaguchi Laboratory, Social ICT Center, University of Tokyo, is investigating this authentication method, and MangaONE is one of candidates for their research.\newline
MangaONE~\cite{mangaone}, one of biggest Manga providers in Japan, with very large number of users and high frequent daily access. Researchers in Yamaguchi Laboratory, Social ICT Center, University of Tokyo, and we are using MangaONE system log information for authentication application, by building a model to distinguish an user's behavior from all others'. Machine Learning, which recently attracts much public attention and many giant corporations' attention, is one of computer science fields be able to be applied on unstructured and multiple features data due to its ability to mimic human like task by computer. There are also many Machine Learning models which can be applied to this authentication of problem. For example, Multilevel Hypergraph Partitioning~\cite{karypis1999multilevel}, Random Forest, Decision tree, k-Nearest Neighbourhood, k-means. R. Kobayashi~\cite{kobayashi1} used Random Forest model to train one model for each user with negative samples randomly chosen from all others. Beside plenty choices of models and acceleration techniques, another topic which data analysts usually care about is programming language. R is one of the most popular language for data analyst in recent years because of its high interactivity, user friendly syntax and flexible data manipulation but its speed is sometimes problematic.\newline
  With machine learning's recent large scale of research community and investment, it has been improved significantly. Basic models with basic algorithms nearly reach their best performance in single machine while input data size keeps increasing with tremendous pace. As a result, there is a need to horizontally scale up the algorithm via GPU acceleration or distributed computing. Horizontally scaling up an application means to divide the application into many smaller tasks in order to execute them in multiple machines in parallel in a network such as distributed system, or execute them with GPUs. According to CRAN Task View~\cite{cran_task}, there are many GPU libraries to speedup R's standard application, such as gputools, gpuR,\ldots. However they almost support only basic matrix and vector manipulation and are lack of Machine Learning model training using GPUs. Gputools is a rare library that trains General Linear model using GPU but it only supports single precision number. R. Kobayashi~\cite{kobayashi1} provided very efficient model to identify user with relatively low false rate but his model requires training one model for each individual. This reduces overall scalability of the solution. His model also ignores majority of input data when training each model because taking all negative data will bias the model and the result predictor will become an one-way negative predictor.\newline
	Our research investigated on using one single model to distinguish any user behaviour in database. Then afterward, we integrated GPU into the standard implementation to speedup the critical part of the training algorithm. In acceleration phase, we chose General Linear model to optimize because of its high correction rate and reasonable resulted predictor. The optimization accelerates the algorithm bottleneck portion around double speed.\newline

@@

First, data was filtered by removing all data of first day and last day because these days' log information may not be completed and their portions are small enough to not affect overall performance. Next, all records of same user on same hour of same day were aggregated into one record with a new label representing number of records, i.e.\ number of the user's reading on specific hour in specific day. Then, week day of the reading was also calculated into integer values between 0 and 6, with Sunday be 0, Monday be 1,\ldots and Saturday be 6. Finally, we came up with data table of 4,532,246 records, each record has 26 data features and 1 label feature as described in Table~\ref{table:db_description}\newline

@@

This data design may raise a question on why chapter id feature was removed. In fact, if chapter was a static feature, in other words, if it never changed by time, it would have been considered to be a component of the model. Moreover, chapters are also grouped into 207 groups by theirs manga with many useful additional data. This has potential to reduce input data size while still prevent losing information on what user read. However, if chapter information is included in the model, by time user might lose their interest in reading some chapters, their reading behaviour with the chapter he/she read will be come variate, and the trained model will be invalidated by time.\newline

@@

Longer bar presents longer time it takes to finish the call. The very bottom bar, glm, presents the most outer call be triggered at command line. Subsequently, 2 eval calls were processed as wrapper functions to generalize the algorithm. Glm.fit is the fitting function that iterates a loop to modify coefficient vector until converged. Each iteration calls a \texttt{.Call} subroutine, which is the most intensive code portion and is written in C. Figure~\ref{fig:wrapping} illustrates the program flow. It also denotes that glm.fit subroutine takes 93.42\% of glm call and \texttt{.Call} subroutine takes the majority 58.13\% running time of glm.fit in all iterations. This implies that optimizing the bottleneck \texttt{.Call} subroutine will be able to speed up the algorithm by at most 2.2 times. Although 2.2 times speed up upper bound is not a much gain, the optimization on this subroutine is mandatory and a basis for subsequent development on eliminating iteration overhead.\newline

@@

\begin{figure}[ht]
  \vspace*{-1.5cm}
  \centering
  \includegraphics[width=\columnwidth,natwidth=1200,natheight=742]{qr_ret.png}
  \caption{Training time of 3 implementation3}\label{fig:qr_ret}
\end{figure}
Figure~\ref{fig:qr_ret} shows that CUSOLVER LS solver C extension run around double (2.12 times) faster than R standard version \texttt{C\_Cdqrls}. Because of the major portion of overhead timing, the overall algorithm can speed up merely 1.27 times over maximal ideal speed up rate of 2.2 times. C extension running time of CUSOLVER LS solver is slightly faster than manual LS solver's one. This can be explained by the fact that input matrix is indeed linearly independent, CUSOLVER LS solver implementation only need to run QR decomposition once without caring about linear independence while manual LS solver version does care about. In CUSOLVER LS solver running time, there exists portion named \texttt{R-C protocol copying overhead}. This is time occupied by memory transfer between R program and C extension because CUSOLVER LS solver program does not use optimized interface while manual LS solver does. This portion elimination can be adopted with not much effort.\newline

@@

\begin{figure}[ht]
  \vspace*{-1.5cm}
  \centering
  \includegraphics[width=\columnwidth,natwidth=1200,natheight=742]{c_narrow.png}
  \caption{Isolated C extension time comparison on 9,064,492 rows by 27 columns matrix}\label{fig:c_narrow}
\end{figure}
Figure~\ref{fig:c_square} and Figure~\ref{fig:c_narrow} emphasize matrix dimension's effect on C extension performance. With input matrix of 3,000 rows by 2,000 columns, GPU version speeded up LS solver by 17.94 times and 79.91 times resepectively. While with input matrix of 9,064,492 rows by 27 columns their speedup rate were merely 1.48 times and 1.65 times respectively. Which implies that narrow matrix does not take more advantage of GPU speedup than square matrix.
\section{Conclusion}
Comparing with R. Kobayashi’s result in~\cite{kobayashi1}. General Linear model in this paper decreases both False Acceptance Rate and False Rejection Rate by 1.14 times (from 12.15\% down to 10.69\%) and 1.3 times (from 14.0\% down to 10.60\%) respectively, although these results slightly depends on model choice and false sample generation method. As discussed in methodology section, graph based approach is very promising. If it is well developed, it will not only be helpful on user classification and distinguishment but also has potential on content recommendation system. For example: user next reading chapter suggestion predictor.\newline
Due to major timing fraction of overhead timing, the optimized version of training algorithm achieved relatively high speedup of 1.27 times under 2.2 times upper bound limit. This speedup can be improved much more with convergence check improvement by reducing almost of memory transfer between GPU device and CPU host in middle iterations. The memory transfer should be done only in the first and the last iteration. Last experiment of the research proved that CUDA speedup effect highly depends on input matrix shape. That is to say, the more features input model has the more speedup this research's optimization gains. It is significant to take notice that both Least Squares solver implementations were not fully optimized. CUSOLVER LS solver requires calling QR decomposition twice due to CUDA Toolkit CUSOLVER library's low abstraction. On the other hand, manual LS solver is nothing more than a simple linear QR decomposition with Householder reflection. Blocked Householder QR algorithm in~\cite{Kerr:2009:QDG:1513895.1513904} is one of the fastest QR implementations executing entirely on the GPU~\cite{Kerr:2009:QDG:1513895.1513904}, with some more investigation on capability with linearly dependent system, it is clearly applicable and gains much speedup. In contrast to GPU optimization technique, it is worthwhile to take attention on multicore optimization by installing and loading appropriate lapack library to exploit multicore power.\newline
The last benchmark result on isolated C extension performance proves GPU accelerated algorithm's effectiveness to wide matrix. That is to say the accelerated algorithm gains more speedup on model with high number of features. Especially, when training model with categorical variables expanded, number of features increases proportionally to categorical variables' domain.

>>>>

In recent years, with the explosion of internet and mobile evolution, internet service is becoming bigger and bigger in terms of system complexity and user volume. Concerning user privacy and credentiality, internet service providers never want to give credential to wrong user. An obvious and canonical method is to use a combination of secret password and unique user identification. However, this becomes a burden for service users to remember their password on each service. There also is a trade-off between remembering convenience and security level when choosing a password. User behaviour of sharing password among multiple services raises the password leaking risk. Recently, there is a new way to give credentiality using Lifestyle Authentication. Lifestyle Authentication is a brand new authentication method which makes use of user auxiliary information rather than user id and password combination~\cite{weko_175884_1}. Auxiliary information can be user location paired with system access timestamp, user daily access pattern, etc\ldots. With this authentication method, a system can identify and authorize a user without requiring him/her doing any procedure explicitly. Yamaguchi Laboratory, Social ICT Center, University of Tokyo, is investigating this authentication method, and MangaONE is one of candidates for their research.\newline
MangaONE~\cite{mangaone}, one of biggest Manga providers in Japan, with very large number of users and high frequent daily access. Researchers in Yamaguchi Laboratory, Social ICT Center, University of Tokyo, and we are using MangaONE system log information for authentication application, by building a model to distinguish an user's behavior from all others'. Machine Learning, which recently attracts much public attention and many giant corporations' attention, is one of computer science fields which can be able to be applied on unstructured and multiple features data due to its ability to mimic human like task by computer. There are also many Machine Learning models which can be applied to this authentication problem. For example, Multilevel Hypergraph Partitioning~\cite{karypis1999multilevel}, Random Forest, Decision tree, k-Nearest Neighbourhood, k-means. R. Kobayashi~\cite{kobayashi1} used Random Forest model to train one model for each user with negative samples randomly chosen from all others. Beside plenty choices of models and acceleration techniques, another topic which data analysts usually care about is programming language. R is one of the most popular language for data analyst in recent years because of its high interactivity, user friendly syntax and flexible data manipulation but its speed is sometimes problematic.\newline
  With machine learning's recent large scale research community and investment, it has been improved significantly. Basic models with basic algorithms nearly reach their best performance in single machine while input data size keeps increasing with tremendous pace. As a result, there is a need to horizontally scale up the algorithm via GPU acceleration or distributed computing. Horizontally scaling up an application means to divide the application into many smaller tasks in order to execute them in multiple machines in parallel in a network such as distributed system, or execute them with GPUs. According to CRAN Task View~\cite{cran_task}, there are many GPU libraries to speedup R's standard application, such as gputools, gpuR,\ldots. However they almost support only basic matrix and vector manipulation and lack Machine Learning model training using GPUs. Gputools is a rare library that trains General Linear model using GPU but it only supports single precision number. R. Kobayashi~\cite{kobayashi1} provided very efficient model to identify user with relatively low false rate but his model requires training one model for each individual. This reduces overall scalability of the solution. His model also ignores majority of input data when training each model because taking all negative data will bias the model and the result predictor will become an one-way negative predictor.\newline
	Our research investigated on using one single model to distinguish any user behaviour in database. Then afterward, we integrated GPU into the standard implementation to speedup the critical part of the training algorithm. In acceleration phase, we chose General Linear model to optimize because of its high correction rate and reasonable resulted predictor. The optimization accelerates the algorithm bottleneck portion around twice.\newline

@@

First, data was filtered by removing all data of first day and last day because these days' log information may not be completed and their portions are small enough to not affect overall performance. Next, all records of same user on same hour of same day were aggregated into one record with a new label representing number of records, i.e.\ number of the user's reading on specific hour in specific day. Then, week day of the reading was also calculated into integer values between 0 and 6, with Sunday be 0, Monday be 1,\ldots and Saturday be 6. Finally, we came up with data table of 4,532,246 records, each record having 26 data features and 1 label feature as described in Table~\ref{table:db_description}\newline

@@

This data design may raise a question on why chapter id feature was removed. In fact, if chapter was a static feature, in other words, if it never changed by time, it would have been considered to be a component of the model. Moreover, chapters are also grouped into 207 groups by theirs manga with many useful additional data. This has potential to reduce input data size while still preventing loss of information on what user read. However, if chapter information is included in the model, by time user might lose his/her interest in reading some chapters, his/her reading behaviour with the chapter he/she read becomes variate, and the trained model will be invalidated.\newline

@@

Longer bar presents longer time it takes to finish the call. The very bottom bar, glm, presents the most outer call be triggered at command line. Subsequently, 2 eval calls were processed as wrapper functions to generalize the algorithm. Glm.fit is the fitting function that iterates a loop to modify coefficient vector until converged. Each iteration calls a \texttt{.Call} subroutine, which is the most intensive code portion and is written in C. Figure~\ref{fig:wrapping} illustrates the program flow. It also denotes that glm.fit subroutine takes 93.42\% of glm call and \texttt{.Call} subroutine takes the majority 58.13\% running time of glm.fit in all iterations. This implies that optimizing the bottleneck \texttt{.Call} subroutine will make it possible to speed up the algorithm by at most 2.2 times. Although 2.2 times speed up upper bound is not a much gain, the optimization on this subroutine is mandatory and a basis for subsequent development on eliminating iteration overhead.\newline

@@

\begin{figure}[ht]
  \vspace*{-1.5cm}
  \centering
  \includegraphics[width=\columnwidth,natwidth=1200,natheight=742]{qr_ret.png}
  \caption{Training time of 3 implementation3}\label{fig:qr_ret}
\end{figure}
Figure~\ref{fig:qr_ret} shows that CUSOLVER LS solver C extension run 2.12 times faster than R standard version \texttt{C\_Cdqrls}. Because of the major portion of overhead timing, the overall algorithm can speed up merely 1.27 times over maximal ideal speed up rate of 2.2 times. C extension running time of CUSOLVER LS solver is slightly faster than manual LS solver's one. This can be explained by the fact that input matrix is indeed linearly independent, CUSOLVER LS solver implementation only need to run QR decomposition once without caring about linear independence while manual LS solver version does care about. In CUSOLVER LS solver running time, there exists portion named \texttt{R-C protocol copying overhead}. This is time occupied by memory transfer between R program and C extension because CUSOLVER LS solver program does not use optimized interface while manual LS solver does. This portion elimination can be adopted with not much effort.\newline

@@

\begin{figure}[ht]
  \vspace*{-1.5cm}
  \centering
  \includegraphics[width=\columnwidth,natwidth=1200,natheight=742]{c_narrow.png}
  \caption{Isolated C extension time comparison on 9,064,492 rows by 27 columns matrix}\label{fig:c_narrow}
\end{figure}
Figure~\ref{fig:c_square} and Figure~\ref{fig:c_narrow} emphasize matrix dimension's effect on C extension performance. With input matrix of 3,000 rows by 2,000 columns, GPU version speeded up LS solver by 17.94 times and 79.91 times resepectively. While with input matrix of 9,064,492 rows by 27 columns their speedup rate were merely 1.48 times and 1.65 times respectively. This implies that narrow matrix does not take more advantage of GPU speedup than square matrix.
\section{Conclusion}
Comparing with R. Kobayashi’s result in~\cite{kobayashi1}. General Linear model in this paper decreases both False Acceptance Rate and False Rejection Rate by 1.14 times (from 12.15\% down to 10.69\%) and 1.3 times (from 14.0\% down to 10.60\%) respectively, although these results slightly depends on model choice and false sample generation method. As discussed in methodology section, graph based approach is very promising. If it is well developed, it will not only be helpful on user classification and distinguishment but also has potential on content recommendation system. For example: user next reading chapter suggestion predictor.\newline
Due to major timing fraction of overhead timing, the optimized version of training algorithm achieved relatively high speedup of 1.27 times under 2.2 times upper bound limit. This speedup can be improved much more with convergence check improvement by reducing almost of memory transfer between GPU device and CPU host in middle iterations. The memory transfer should be done only in the first and the last iteration. Last experiment of the research proved that CUDA speedup effect highly depends on input matrix shape. That is to say, the more features input model has the more speedup this research's optimization gains. It is significant to take notice that both Least Squares solver implementations were not fully optimized. CUSOLVER LS solver requires calling QR decomposition twice due to CUDA Toolkit CUSOLVER library's low abstraction. On the other hand, manual LS solver is nothing more than a simple linear QR decomposition with Householder reflection. Blocked Householder QR algorithm in~\cite{Kerr:2009:QDG:1513895.1513904} is one of the fastest QR implementations executing entirely on the GPU~\cite{Kerr:2009:QDG:1513895.1513904}, with some more investigation on capability with linearly dependent system, it is clearly applicable and gains much speedup. In contrast to GPU optimization technique, it is worthwhile to take attention on multicore optimization by installing and loading appropriate lapack library to exploit multicore power.\newline
The last benchmark result on isolated C extension performance proves GPU accelerated algorithm's effectiveness to wide matrix. That is to say the accelerated algorithm gains more speedup on model with high number of features. Especially, when training model with categorical variables expanded, number of features increases proportionally to categorical variables' domain.
